---
title: "Portfolio1"
author: "Ingrid Backman"
date: "2023-03-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
set.seed(123)
```

1. _(5.2 from ROS)_ __Continuous probability simulation:__ The logarithms of weights (in pounds) of men in the United States are approximately normally distributed with mean 5.13 and standard deviation 0.17; women’s log weights are approximately normally distributed with mean 4.96 and standard deviation 0.20. Suppose 10 adults selected at random step on an elevator with a capacity of 1750 pounds. What is the probability that their total weight exceeds this limit?
```{r}
#these are all in log
m_mean <- 5.13
m_sd <- 0.17
w_mean <- 4.96
w_sd <- 0.20

#these are not in log
limit <- 1750
n_adults <- 10

#10 adults are chosen at random to step on an elevator with a capacity of 1750 pounds, what are the odds that their weight exceeds the limit?

#what are the odds of getting a male and a woman at random in USA? according to Google 50.5% of USA is female
f_odds <- 0.505
m_odds <- 1-f_odds

n <- 1000

#initialize the total_weight before loop
total_weight <- rep(NA, n)

for (i in 1:n) {
    male <- rbinom(n_adults, 1, m_odds) #generating the gender ratio by using the male odds, so male is 1
    
    log_weight <- ifelse(
        male == 1,
        rnorm(n_adults, m_mean, m_sd),
        rnorm(n_adults, w_mean, w_sd)) #generating the weights with mean, sd, male = 1, female = 0
    
    #exp() is the opposite of log! need to convert since the weight limit isn't in log!
    weight <- exp(log_weight)
    total_weight[i] <- sum(weight)
}

exceeded_limit <- sum(total_weight > limit)
percentage_exceeded_limit <- exceeded_limit / n * 100
cat("The total weight exceeded the limit in:", percentage_exceeded_limit, "% of trials")

hist(total_weight)
  abline(v = 1750, col = "red")
```

2. _(5.6 from ROS)_ __Propagation of uncertainty:__ We use a highly idealized setting to illustrate the use of simulations in combining uncertainties. Suppose a company changes its technology for widget production, and a study estimates the cost savings at \$5 per unit, but with a standard error of \$4. Furthermore, a forecast estimates the size of the market (that is, the number of widgets that will be sold) at 40 000, with a standard error of 10 000. Assuming these two sources of uncertainty are independent, use simulation to estimate the total amount of money saved by the new product (that is, savings per unit, multiplied by size of the market).
$${\displaystyle \operatorname {Var} (XY)=(\sigma _{X}^{2}+\mu _{X}^{2})(\sigma _{Y}^{2}+\mu _{Y}^{2})-\mu _{X}^{2}\mu _{Y}^{2}}$$
When two random variables are statistically independent, the expectation of their product is the product of their expectations.

```{r}
#the equivalents of mean and sd
save_per_unit <- 5
unit_sd <- 4
market <- 40000
market_sd <- 10000
#savings per unit, multiplied by size of the market
#When two random variables are statistically independent, the expectation of their product is the product of their expectations. --> multiply

n <- 10000 #do i simply choose the amount of n?

savings_samples <- rnorm(n, save_per_unit, unit_sd)
market_samples <- rnorm(n, market, market_sd)

total_savings <- savings_samples * market_samples

mean_savings <- mean(total_savings)

cat("Estimated total amount of money saved:", mean_savings, "dollars")

hist(savings_samples)
hist(market_samples)
hist(total_savings)
```
3. _(5.10 from ROS)_ __Inference for a ratio of parameters:__ A (hypothetical) study compares the costs and effectiveness of two different medical treatments.

    - In the first part of the study, the difference in costs between treatments A and B is estimated at \$600 per patient, with a standard error of \$400, based on a regression with 50 degrees of freedom.
    - In the second part of the study, the difference in effectiveness is estimated at 3.0 (on some relevant measure), with a standard error of 1.0, based on a regression with 100 degrees of freedom.
    - For simplicity, assume that the data from the two parts of the study were collected independently.

    Inference is desired for the incremental cost-effectiveness ratio: the difference between the average costs of the two treatments, divided by the difference between their average effectiveness, a problem discussed further by Heitjan, Moskowitz, and Whang (1999).

    (a) Create 1000 simulation draws of the cost difference and the effectiveness difference, and make a scatterplot of these draws.
```{r}
#1st part of the study
cost_diff <- 600
cost_se <- 400
cost_df <- 50

#2nd part of the study
de <- 3.0
de_se <- 1.0
de_df <- 100

#Inference is desired for the incremental cost-effectiveness ratio: the difference between the average costs of the two treatments, divided by the difference between their average effectiveness

n <- 1000

#using the degrees of freedom to get standard deviation from standard error
#SE = SD / sqrt(n) 
#SD = SE * sqrt(n)
cost_sd <- cost_se * sqrt(cost_df)
de_sd <- de_se * sqrt(de_df)


diff_cost <- rnorm(n, cost_diff, cost_sd) #difference cost
diff_eff <- rnorm(n, de, de_sd) #difference effect


plot(diff_cost, diff_eff, xlab = "Cost Difference", ylab = "Effectiveness Difference")

```

(b) Use simulation to come up with an estimate, 50% interval, and 95% interval for the incremental cost-effectiveness ratio.
```{r}
ratio <- diff_cost/diff_eff
estimate <- mean(ratio)

c_50 <- quantile(ratio, c(.25, .75))
c_95 <- quantile(ratio, c(.05, .95))

cat("estimate: ", estimate, "\n", "50% confidence interval: from 25%:", c_50[1],"to 75%:",c_50[2], "\n", "95% confidence interval: from 5%:", c_95[1],"to 95%:",c_95[2])
```
(c) Repeat, changing the standard error on the difference in effectiveness to 2.0.
```{r}
#only change
de_se <- 2.0

#just a redo
cost_sd <- cost_se * sqrt(cost_df)
de_sd <- de_se * sqrt(de_df)


diff_cost <- rnorm(n, cost_diff, cost_sd) #difference cost
diff_eff <- rnorm(n, de, de_sd) #difference effect


plot(diff_cost, diff_eff, xlab = "Cost Difference", ylab = "Effectiveness Difference")

ratio <- diff_cost/diff_eff
estimate <- mean(ratio)

c_50 <- quantile(ratio, c(.25, .75))
c_95 <- quantile(ratio, c(.05, .95))

cat("estimate: ", estimate, "\n", "50% confidence interval: from 25%:", c_50[1],"to 75%:",c_50[2], "\n", "95% confidence interval: from 5%:", c_95[1],"to 95%:",c_95[2])
```

