---
title: "R Notebook"
output:
  html_document:
    df_print: paged
---

# Exercises from Chapter 3
## Warm-up: Practicing vector and matrix notation
Consider the following predictions
$$
\begin{aligned}
& \hat{y}_1=47=50+3.0 *(-1) \\
& \hat{y}_2=-50=50*(-1)+3.0 * 0 \\
& \hat{y}_3=109=50*2+3.0 * 3
\end{aligned}
$$
Try to write the prediction in vector notation. If you further want to challenge yourself try Matrix notation as well. (if you are stuck have a look at pages 36-37 in ROS). It might be helpful to do this exercise by hand.


## 3.1
Weighted averages: A survey is conducted in a certain city regarding support for increased
property taxes to fund schools. In this survey, higher taxes are supported by 50% of respondents aged 18–29, 60% of respondents aged 30–44, 40% of respondents aged 45–64, and 30% of respondents aged 65 and up. Assume there is no nonresponse.
Suppose the sample includes 200 respondents aged 18–29, 250 aged 30–44, 300 aged 45–64,
and 250 aged 65+. Use the weighted average formula to compute the proportion of respondents
in the sample who support higher taxes.

Higher property taxes support age (18-29) = 50%, n = 200
Higher property taxes support age (30-44) = 60%, n = 250
Higher property taxes support age (45-64) = 40%, n = 300
Higher property taxes support age (65+) = 30%, n = 250

```{r}
library(tidyverse)
library(tibble)
```


```{r}
n <- (c(200, 250, 300, 250, 200))
n_procent <- (c(0.50, 0.60, 0.40, 0.30, 0.20))
age <- (c("18-29", "30-44", "45-64", "65-72", "73-inf"))
df <- tibble(age , n, n_procent)
```


```{r}
w_avg <- function(n, n_procent){
  # making the total variable
   total <- 0
  
  # making a for loop
  for (i in 1:length(n)){
    addition <- n[i]/sum(n)*n_procent[i]
    
    total <- total + addition
    
  }
   total <- total * 100
   
  print(paste(round(total, 2),"%", sep = ""))
}

w_avg(n, n_procent)
```


---

```{r}
n <- (c(200, 250, 300, 250))
n_procent <- (c(0.50, 0.60, 0.40, 0.30))
age <- (c("18-29", "30-44", "45-64", "65+"))
df <- tibble(age , n, n_procent)
```


```{r}
# Making function for weighted average
weighted_average <- function(n, n_procent, age){ for(k in n){
  k/sum(n)*
  print(k)
}
  n[1]/sum(n)*n_procent[1]
}

weighted_average(n, n_procent, age)

weighted_average <- (n[1]/sum(n))*n_procent[1]+(n[2]/sum(n))*n_procent[2]+(n[3]/sum(n))*n_procent[3]+(n[4]/sum(n)*n_procent[4])
```


## 3.3
Probability distributions: Using R, graph probability densities for the normal distribution,
plotting several different curves corresponding to different choices of mean and standard
deviation parameters.

```{r}
mean_n <- seq(-4, 4, by = 0.5) 
sd_n <- seq(0, 4, by = 0.5) 


x <- seq(-6, 6, length.out = 200)
y <- dnorm(x, mean = 0, sd = 1)

# making plot

plot(x, y, type = "l", xlab = "x", ylab = "Probability Density")


for (i in 1:8){
  mean_n <- seq(-4, 4, by = 0.5) 
  sd_n <- seq(1, 4, by = 0.5)
  
  x <- seq(-6, 6, length.out = 200)
  y <- dnorm(x, mean = mean_n[i], sd = sd_n[i])
  
  lines(x, y)
}

```


## 3.4
Probability distributions: Using a bar plot in R, graph the Poisson distribution with parameter
3.5.


```{r}

# Generate a vector of possible values for the Poisson random variable
x <- 0:20

# Compute the probabilities of each possible value using the dpois() function
y <- dpois(x, lambda = 5)

# Display the probabilities as a bar plot using barplot()
barplot(y, names.arg = x, xlab = "Possible Values", ylab = "Probabilities", main = "Poisson Distribution with lambda = 3.5")

```


## 3.5
Probability distributions: Using a bar plot in R, graph the binomial distribution with n = 20 and
p = 0.3.


```{r}
y <- dbinom(0:10, size = 20, prob = 0.3)
x <- 0:10

# Making barplot

barplot(y, names.arg = x, ylim = c(0, 0.25))
```


## 3.8
 Correlated random variables: Suppose that the heights of husbands and wives have a correlation
of 0.3, husbands’ heights have a distribution with mean 69.1 and standard deviation 2.9 inches,
and wives’ heights have mean 63.7 and standard deviation 2.7 inches. Let x and y be the heights
of a married couple chosen at random. What are the mean and standard deviation of the average
height, (x + y)/2?




```{r}
# Generating data for men

set.seed(123)  # set the seed for reproducibility
n <- 10000    # number of values to generate
mu <- 69.1       # desired mean
sigma <- 2.9     # desired standard deviation

x <- rnorm(n, mean = mu, sd = sigma)

# Generating data for Women

set.seed(222)  # set the seed for reproducibility
n <- 10000       # number of values to generate
mu <- 63.7      # desired mean
sigma <- 2.7     # desired standard deviation

y <- rnorm(n, mean = mu, sd = sigma)

df <- tibble(y,x)

cor(x,y)
```
```{r}
# Generating data for men
set.seed(123)  # set the seed for reproducibility
n <- 10000    # number of values to generate
mu <- 69.1       # desired mean
sigma <- 2.9     # desired standard deviation

x <- rnorm(n, mean = mu, sd = sigma)

# Generating data for women
set.seed(123)  # set the seed for reproducibility
n <- 10000    # number of values to generate
mu <- 63.7       # desired mean
sigma <- 2.7     # desired standard deviation

y <- rnorm(n, mean = mu, sd = sigma)




```


# Exercises from Chapter 4
These exercises are optional if you need some extra challenge. It is a sneakpeak of the chapter for next weeks lecture.

## 4.1 

```{r}

```


## 4.2

```{r}

```

