---
title: "R Notebook"
output: html_notebook
---

# By hand or Latex syntax in a .Rmd file?
Most of these exercises are easiest solved by hand. However, if you want to write math notation in an R-markdown Latex-syntax can be used. Below are some examples of how latex is used to write math equations. Remember to surround your equation by either $ $ or $$ $$. Feel free to ask for me if you want to learn some latex syntax. I can also highly recommend mathpix snipping tool for this purpose :)

$\frac{5}{10}$

$$
x^{\frac{a}{b}}=(x^a)^{\frac{1}{b}}=(x^{\frac{1}{b}})^a=\sqrt[b]{x^a}
$$

# 1.1

# 1.2

# 1.3

# 1.4
```{r}
# Simple graphing can go here..
# Define the function
my_function <- function(f) {
  c <- 5/9 * f - 160/9
  return(c)
}

# Create a sequence of values for f
f_seq <- seq(-40, 140, by = 1)

# Calculate the corresponding values of c
c_seq <- my_function(f_seq)

# Plot the graph
plot(f_seq, c_seq, type = "l", xlab = "Fahrenheit", ylab = "Celsius")

```

# 1.6

# 1.15

# 1.21
Expected effect is here the expected change in homicides per 100,000

# 1.22
Figure out what the modulo operator does and solve the exercise

The modulo operator is simply implemented in R by %%

```{r}
# Example
5 %% 2 
```





