---
title: "Week8"
author: "Christian Vejen Barnkob"
date: "2023-03-31"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

6.4
For the function $$f(x,y)=\frac{sin(xy)}{cos(x+y)}$$ calculate the partial derivatives
with respect to x and y.

$$\frac{\partial\ }{\partial x}\frac{sin(xy)}{cos(x+y)}=\frac{cos(x+y)\cdot cos(xy)\cdot y-sin(xy)\cdot (-sin(x+y)\cdot1)}{cos(x+y)^2}$$


$$\frac{\partial }{\partial y}\frac{sin(xy)}{cos(x+y)}=\frac{cos(x+y)\cdot cos(xy)\cdot x-sin(xy)\cdot (-sin(x+y))}{cos(x+y)^2}$$


6.7
For the function $$f(u,v)=\sqrt{u+v^2}$$ calculate the partial derivatives
with respect to $u$ and $v$ and provide the value of these functions at the
point 
$$f(u,v)=\sqrt{u+v^2}$$
$$(\frac{1}{2}, \frac{1}{3})$$

$$\frac{\partial }{\partial u}\sqrt{u+v^2}=\frac{1}{2}(u+v^2)^{-\frac{1}{2}}\cdot 1$$ 
$$=\frac{1}{2}(\frac{1}{2}+\frac{1}{3}{}^2)^{-\frac{1}{2}}\cdot 1=\frac{3}{22} \sqrt{22}$$ 
6.9

Done in hand

6.1 (only do the first 4 and try to graph the functions in R)


```{r}
x <- seq(-10, 10, length.out = 1000)

y <- 1/x

plot(x, y, type = "l", xlab = "x", ylab = "y", main = "Plot of 1/x")
```

```{r}
x <- seq(-10, 10, length.out = 1000)

y <- x^3

plot(x, y, type = "l", xlab = "x", ylab = "y", main = "Plot of x^3")
```

```{r}
x <- seq(-10, 10, length.out = 1000)

y <- x^2+4*x+8

plot(x, y, type = "l", xlab = "x", ylab = "y", main = "Plot of x^3") 
```

If you have more time. Finish exercises from last week or do more exercises from chapter 6 (for instance 6.11).

